### 2020-06-02 writeup/intro.tex
intro wip

In this work, we consider the setting of high-dimensional linear regression and focus on predicting a particular task that only contains limited amount of labeled data.
Following Hastie et al. \cite{HMRT19} and Bartlett et al. \cite{BLLT20}, we assume that for every task $1\le i\le t$, its features are random vectors $x = \Sigma_i^{1/2}z$, where $z\in\real^p$ consists of i.i.d. entries with mean zero and unit variance, and $\Sigma_i\in\real^{p\times p}$ is a positive semidefinite matrix.
Let $n_i$ denote the data size and $X_i\in\real^{n_i\times p}$ denote the features of task $i$.
The labels of task $i$ are given by $Y_i = X_i\beta_i + \varepsilon_i$, where $\beta_i\in\real^p$ denotes task $i$'s ground truth parameters and $\varepsilon_i$ denotes i.i.d. random noise with mean zero and variance $\sigma^2$.
Without loss of generality, let the $t$-th task denote the target task.
Importantly, the target task's data size is a fixed constant times $p$ in the high-dimensional setting.
Hence adding more labeled data can help improve its test performance.
%We shall assume that each task data follows a linear model, i.e. $y_i = X_i \beta_i + \varepsilon_i$, $1\le i\le k$.
%Here $\beta_i\in\real^p$ is the model parameter for the $i$-th task.
%Each row of $X_i\in\real^{n_i\times p}$ is assumed to be drawn i.i.d. from a fixed
%distribution with covariance matrix $\Sigma_i$.
We focus on the hard parameter sharing architecture with a linear shared layer $B\in\real^{p\times r}$ for all tasks and a separate prediction head $\set{W_i \in \real^{r}}_{i=1}^t$ for each task \cite{R17,MTDNN19,WZR20}.
%This corresponds to minimizing $ \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2$.
Let $\hat{\beta}_t^{\MTL} = B W_t$ denote the optimal multi-task estimator for the target task.
Let $\hat{\beta}_t^{\STL}$ denote the single-task estimator.
We define these precisely in Section \ref{sec_setup}.
We say there is negative transfer if the prediction loss of $\hat{\beta}_t^{\MTL}$  is larger than that of $\hat{\beta}_t^{\STL}$, or positive transfer otherwise.

@@

\textbf{Main results.}
We revisit the bias-variance tradeoff of the multi-task estimator.
Interestingly, we observe that the variance of the multi-task estimator is always smaller than the variance of the single-task estimator, hence resulting in a positive effect of variance reduction.
The bias of the multi-task estimator results in a negative effect caused by the difference between $\beta_t$ and the rest of $\set{\beta_i}_{i=1}^{t-1}$.
Hence, the bias-variance tradeoff determines whether we observe positive or negative transfer.

Our first contribution is to develop a new technical tool for a quantity that arises naturally from the analysis of two tasks.
Concretely, we show a tight bound on the trace of $(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}$, which extends a well-known result on the trace of $(X_1^{\top}X_1)^{-1}$ for the single-task setting \cite{S07}.
Our analysis yields nearly optimal concentration error, which may be of independent interest.
Using this tool, we show the following results in Section \ref{sec_main}.
\squishlist
	\item We provide a sharp analysis of the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$ for two settings:
(i) two tasks with general covariance matrice; (ii) any number of tasks that have the same features (e.g. images).
	\item We extend our result to transfer learning in the setting of high-dimensional linear regression.
	We pool learnt source task representations into a shared body similar to $B$ in the hard parameter sharing architecture, as in Taskonomy by Zamir et al. \cite{ZSSGM18}.
	We show that the model shift bias can be captured by the projection of $\beta_t$ to the orthogonal subspace spanned by $\set{\beta_i}_{i=1}^{t-1}$.
\squishend

Our second contribution is to use our newly developed tool to explain negative transfer in multi-task learning rigorously.
In Section \ref{sec_insight}, we consider three components including task similarity, data size and covariate shift for a simplified isotropic setting of two tasks.
We measure task similarity by how small is the distance between $\beta_1$ and $\beta_2$.
Using our tool, we explain a transition from positive to negative transfer as task similarity decreases.
%		Furthermore, we show that negative transfer is more likely to occur when the source task labels are particularly noisy.
%		In Section \ref{sec_validate}, we validate the observation on text and image classification tasks.
%	In , we provide the trade-off between $\norm{\beta_1 - \beta_2}^2$ and a certain function $\Phi(\rho_1, \rho_2)$ to determine the type of transfer.
We show that increasing the data size of the source task does not always improve performance for the target task in multi-task learning.
	Then, we analyze the benefit of multi-task learning for reducing the amount of labeled data needed to achieve comparable performance to single-task learning, which is a key empirical finding of Taskonomy.
We show that covariate shift, measured by $\Sigma_1^{1/2}\Sigma_2^{-1/2}$, is another cause for suboptimal performance for $\hat{\beta}_t^{\MTL}$.
		We show that as $n_1 / n_2$ becomes large, having no covariate shift between the source and target tasks yields the optimal performance for the target task.
%		On the other hand, when $n_1 / n_2$ is small, there are counter examples where having the same covariance matrix is not necessarily the optimal choice.

>>>>

\textbf{Setup and Main results.}
In this work, we consider the setting of high-dimensional linear regression \cite{HMRT19,BLLT20} and focus on predicting a particular task that only contains limited amount of labeled data.
%Following Hastie et al.  and Bartlett et al. \cite{},
We assume that for every task $1\le i\le t$, its features are random vectors $x = \Sigma_i^{1/2}z$, where $z\in\real^p$ consists of i.i.d. entries with mean zero and unit variance, and $\Sigma_i\in\real^{p\times p}$ is a positive semidefinite matrix.
Let $n_i = \rho_i \times p$ denote the data size and $X_i\in\real^{n_i\times p}$ denote the features of task $i$.
The labels of task $i$ are given by $Y_i = X_i\beta_i + \varepsilon_i$, where $\beta_i\in\real^p$ denotes task $i$'s ground truth parameters and $\varepsilon_i$ denotes i.i.d. random noise with mean zero and variance $\sigma^2$.
We focus on the hard parameter sharing architecture with a linear shared layer for all tasks and a separate prediction head for each task \cite{R17,MTDNN19,WZR20}.
%This corresponds to minimizing $ \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2$.
Let $\hat{\beta}_t^{\MTL} = B W_t$ denote the optimal multi-task estimator for the target task, which is defined precisely in Section \ref{sec_setup}.
Our main theoretical contribution is to develop a new technical tool for a quantity that arises naturally from the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$ for two tasks.
%Let $\hat{\beta}_t^{\STL}$ denote the single-task estimator.
%Without loss of generality, let the $t$-th task denote the target task.
%Importantly, the target task's data size is a fixed constant times $p$ in the high-dimensional setting.
%Hence adding more labeled data can help improve its test performance.
%$B\in\real^{p\times r}$
%$\set{W_i \in \real^{r}}_{i=1}^t$

@@

%We revisit the bias-variance tradeoff of the multi-task estimator.
%Interestingly, we observe that the variance of the multi-task estimator is always smaller than the variance of the single-task estimator, hence resulting in a positive effect of variance reduction.
%The bias of the multi-task estimator results in a negative effect caused by the difference between $\beta_t$ and the rest of $\set{\beta_i}_{i=1}^{t-1}$.
%Hence, the bias-variance tradeoff determines whether we observe positive or negative transfer.

\begin{theorem}[Informal]\label{lem_cov_shift_informal}
	In the setting of two tasks, let $a_1, a_2$ be fixed values solved from deterministic equations given by $\rho_1, \rho_2$ and the singular values of $\Sigma_1^{1/2}\Sigma_2^{-1}\Sigma_1^{1/2}$.
	%let $M = \Sigma_1^{1/2}\Sigma_2^{-1/2}$ and $\lambda_1, \lambda_2, \dots, \lambda_p$ be the singular values of $M^{\top}M$ in descending order.
	For any constant $\e>0$, w.h.p. over the randomness of $X_1, X_2$, we have that
	\begin{align*}
		\bigtr{(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2} = \frac{1}{\rho_1+\rho_2}\cdot \frac1p\bigtr{ (a_1 \Sigma_1 + a_2\Sigma_2)^{-1} \Sigma_2} +\bigo{\frac{\|\Sigma_2\|}{p^{1/2-\epsilon}}}.
	\end{align*}
\end{theorem}

%Concretely, we show a tight bound on the trace of $(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}$, which
Theorem \ref{lem_cov_shift_informal} extends a well-known result for the single-task setting \cite{S07}, e.g. when $X_1 = 0$.
The concentration error of our result, which is on the order of $O(p^{-1/2+\varepsilon})$, is nearly optimal.
Using this tool,
%We shall assume that each task data follows a linear model, i.e. $y_i = X_i \beta_i + \varepsilon_i$, $1\le i\le k$.
%Here $\beta_i\in\real^p$ is the model parameter for the $i$-th task.
%Each row of $X_i\in\real^{n_i\times p}$ is assumed to be drawn i.i.d. from a fixed
%distribution with covariance matrix $\Sigma_i$.
%We say there is negative transfer if the prediction loss of $\hat{\beta}_t^{\MTL}$  is larger than that of $\hat{\beta}_t^{\STL}$, or positive transfer otherwise.
we provide a sharp analysis of the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$ for two settings:
(i) two tasks with general covariance matrice; (ii) any number of tasks that have the same features (e.g. images).
We extend our result to transfer learning in the setting of high-dimensional linear regression.
We pool learnt source task representations into the shared body of the hard parameter sharing architecture, as in Taskonomy by Zamir et al. \cite{ZSSGM18}.
We show that the bias of the transfer learning estimator can be captured by the projection of $\beta_t$ to the orthogonal subspace spanned by $\set{\beta_i}_{i=1}^{t-1}$.
These results are described more precisely in Section \ref{sec_main}.

Our second contribution is to use our newly developed tool to explain negative transfer in multi-task learning rigorously.
In Section \ref{sec_insight}, we explain the three phenomena observed in Figure \ref{fig_model_shift_phasetrans} for a simplified isotropic setting of two tasks.
%In Section \ref{sec_insight}, we consider three components including task similarity, data size and covariate shift for a simplified isotropic setting of two tasks.
%We measure task similarity by how small is the distance between $\beta_1$ and $\beta_2$.
%Using our tool, we explain a transition from positive to negative transfer as task similarity decreases.
%		Furthermore, we show that negative transfer is more likely to occur when the source task labels are particularly noisy.
%		In Section \ref{sec_validate}, we validate the observation on text and image classification tasks.
%	In , we provide the trade-off between $\norm{\beta_1 - \beta_2}^2$ and a certain function $\Phi(\rho_1, \rho_2)$ to determine the type of transfer.
%We show that increasing the data size of the source task does not always improve performance for the target task in multi-task learning.
Then, we analyze the benefit of multi-task learning for reducing the amount of labeled data needed to achieve comparable performance to single-task learning, which is a key empirical finding of Taskonomy.
%We show that covariate shift, measured by $\Sigma_1^{1/2}\Sigma_2^{-1/2}$, is another cause for suboptimal performance for $\hat{\beta}_t^{\MTL}$.
%		We show that as $n_1 / n_2$ becomes large, having no covariate shift between the source and target tasks yields the optimal performance for the target task.
%		On the other hand, when $n_1 / n_2$ is small, there are counter examples where having the same covariance matrix is not necessarily the optimal choice.

### 2020-06-02 writeup/setup.tex
sec 2
Sec A
intro wip

\begin{lemma}[Informal statement of Lemma \ref{lem_cov_shift}]\label{lem_cov_shift_informal}
	In the setting of two tasks, let $M = \Sigma_1^{1/2}\Sigma_2^{-1/2}$ and $\lambda_1, \lambda_2, \dots, \lambda_p$ be the singular values of $M^{\top}M$ in descending order.
	For any constant $\e>0$, w.h.p. over the randomness of $X_1, X_2$, we have that
	\begin{align*}
		\bigtr{(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2} = \frac{1}{\rho_1+\rho_2}\cdot \frac1p\bigtr{ (a_1 \Sigma_1 + a_2\Sigma_2)^{-1} \Sigma_2} +\bigo{\|\Sigma_2\| p^{-1/2+\epsilon}},
	\end{align*}
where $(a_1, a_2)$ is the solution to the following deterministic equations:
	\begin{align*}
		a_1 + a_2 = 1- \frac{1}{\rho_1 + \rho_2},\quad a_1 + \frac1{\rho_1 + \rho_2}\cdot \frac{1}{p}\sum_{i=1}^p \frac{\lambda_i^2 a_1}{\lambda_i^2 a_1 + a_2} = \frac{\rho_1}{\rho_1 + \rho_2}.
	\end{align*}
\end{lemma}

@@

\begin{theorem}[Informal statement of Theorem \ref{thm_model_shift}]\label{thm_main_informal}
	For the setting of two tasks, let $M=\Sigma_1^{1/2}\Sigma_2^{-1/2}$, $\delta > 0$ be a desired error margin and $\rho_1 \gtrsim \frac{1}{\delta^2}\cdot \lambda_{\min}(M)^{-4} \norm{\Sigma_1} \max(\norm{\beta_1}^2, \norm{\beta_2}^2)$, where $\lambda_{\min}(M)$ is the smallest singular value of $M_1$.
	There exists two deterministic functions $\Delta_{\bias}$ and $\Delta_{\vari}$ that only depend on $\set{\hat{v}, \Sigma_1, \Sigma_2, \rho_1, \rho_2, \beta_1, \beta_2}$ such that
	\squishlist
		\item If $\Delta_{\bias} + \Delta_{\vari} < -\delta$, then w.h.p. over the randomness of $X_1, X_2$, we have $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\Delta_{\bias} - \Delta_{\vari} > \delta$, then w.h.p. over the randomness of $X_1, X_2$, we have $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\squishend
\end{theorem}

@@

\textbf{Transfer learning.}
We extend Theorem \ref{thm_main_informal} to transfer learning settings.
We study the transfer procedure used in Taskonomy by Zamir et al.'18 \cite{ZSSGM18}.
For the setting of high-dimensional linear regression, the transfer procedure is as follows.
%Specifically, the source task encoder consists of the representations learnt from one or more source tasks.
%The transfer function then tries to fit the target task data to the source task encoder.
\squishlist
	\item \textit{Learning source task representations}: we obtain $\hat{\beta}_i^{\STL}$ from each source task, for $1\le i \le t-1$.
		This forms a shared representation $\hat{B} = [\hat{\beta}_1,\hat{\beta}_2,\dots,\hat{\beta}_{t-1}]$.
	\item \textit{Fine-tuning on the target task}: we learn the output layer $W_t\in\real^{t-1}$ similar to equation \eqref{eq_mtl}
		\begin{align}
			g(W_t) = \bignorm{X_t B W_t - Y_t}^2.
		\end{align}
\squishend
After solving $W_t$, we use $\hat{\beta}_t^{\TL} = B W_t$ as the transfer learning (TL) estimator for the target task.
Our result naturally applies to $\hat{\beta}_t^{\TL}$.
Interestingly, we show that the model shift bias is simply the projection of $\beta_t$ to the orthogonal subspace of $B^{\star} = [\beta_1,\dots,\beta_{t-1}]$, or $\norm{\Sigma_t^{1/2}(\id - B^{\star}({B^{\star}}^{\top}B^{\star})^{-1}{B^{\star}}^{\top})\beta_t}^2$ more precisely.
The formal statement is presented in Theorem \ref{prop_taskonomy} and its proof in Appendix \ref{app_proof_sec4}.

>>>>

where $(a_1, a_2)$ is the solution to the following deterministic equations:
	\begin{align*}
		a_1 + a_2 = 1- \frac{1}{\rho_1 + \rho_2},\quad a_1 + \frac1{\rho_1 + \rho_2}\cdot \frac{1}{p}\sum_{i=1}^p \frac{\lambda_i^2 a_1}{\lambda_i^2 a_1 + a_2} = \frac{\rho_1}{\rho_1 + \rho_2}.
	\end{align*}

@@

\begin{theorem}[Informal statement of Theorem \ref{thm_model_shift}]\label{thm_main_informal}
	For the setting of two tasks, let $M=\Sigma_1^{1/2}\Sigma_2^{-1/2}$, $\delta > 0$ be a desired error margin and $\rho_1 \gtrsim \frac{1}{\delta^2}\cdot \lambda_{\min}(M)^{-4} \norm{\Sigma_1} \max(\norm{\beta_1}^2, \norm{\beta_2}^2)$, where $\lambda_{\min}(M)$ is the smallest singular value of $M_1$.
	There exists two deterministic functions $\Delta_{\bias}$ and $\Delta_{\vari}$ that only depend on $\set{\hat{v}, \Sigma_1, \Sigma_2, \rho_1, \rho_2, \beta_1, \beta_2}$ such that
	\squishlist
		\item If $\Delta_{\bias} - \Delta_{\vari} < -\delta$, then w.h.p. over the randomness of $X_1, X_2$, we have $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\Delta_{\bias} - \Delta_{\vari} > \delta$, then w.h.p. over the randomness of $X_1, X_2$, we have $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\squishend
\end{theorem}

@@

\textbf{Transfer learning.}
We extend Theorem \ref{thm_main_informal} to transfer learning settings.
We study the transfer procedure used in Taskonomy by Zamir et al.'18 \cite{ZSSGM18}.
For the setting of high-dimensional linear regression, the transfer procedure is as follows.
%Specifically, the source task encoder consists of the representations learnt from one or more source tasks.
%The transfer function then tries to fit the target task data to the source task encoder.
\squishlist
	\item \textit{Learning source task representations}: we obtain $\hat{\beta}_i^{\STL}$ from each source task, for $1\le i \le t-1$.
		This forms a shared representation ${B} = [\hat{\beta}_1,\hat{\beta}_2,\dots,\hat{\beta}_{t-1}]$.
	\item \textit{Fine-tuning on the target task}: we learn the output layer $W_t\in\real^{t-1}$ similar to equation \eqref{eq_mtl}
		\begin{align}
			g(W_t) = \bignorm{X_t B W_t - Y_t}^2.
		\end{align}
\squishend
After solving $W_t$, we use $\hat{\beta}_t^{\TL} = B W_t$ as the transfer learning (TL) estimator for the target task.
Our result naturally applies to $\hat{\beta}_t^{\TL}$.
Interestingly, we show that the model shift bias is simply the projection of $\beta_t$ to the orthogonal subspace of $B^{\star} = [\beta_1,\dots,\beta_{t-1}]$, or $\norm{\Sigma_t^{1/2}(\id - B^{\star}({B^{\star}}^{\top}B^{\star})^{-1}{B^{\star}}^{\top})\beta_t}^2$ more precisely.
The formal statement is presented in Theorem \ref{prop_taskonomy} and its proof in Appendix \ref{app_proof_sec4}.
